---
title: "Indo-European Pronouns"
output:
  pdf_document: default
  html_notebook: default
  html_document:
    df_print: paged
---

This file contains all the source code to generate the figures and statistical measures reported in the paper **Evolution and Spread of Politeness Systems in Indo-European**

Required data files

- `dollo.combined.trees` (a sample of bayesian phylogenetic trees for the I-E language family)
- `dollo.mcc.tre` (a Maximum Clade Consistency tree summarising the tree sample, for reference only)
- `language.metadata.csv` (ISO codes, geographic coordinates, and ancient/modern status)
- `SAE-features.csv` (presence/absence of SAE areal features, coded from Haspelmath list)
- `../data/pronoun-paradigms.csv` (an etic grid showing the pronominal forms used for each pronominal function attested in the family)

Generated files:

- `pronoun-similarity.csv` (a pairwise similarity table, generated from `pronoun-paradigms.csv` using the python script `../data/make-pronoun-similarity-table.py`; for each language each pair of pronominal functions are rated 1/same pronoun or 0/different pronouns)
- `pronoun-number-conflation.csv` (generated from `pronoun-paradigms.csv` with the script `../data/make_pronoun_conflation_table.py`; does the language ever conflate singlular and plural for 2nd or 3rd person)
- `trees.rData` (saved intermediate results of this analysis)
- `simmap.rData` (saved intermediate results of this analysis)
- `results.rData` (saved results of this analysis)

To regenerate the analysis cleanly you need to delete the three `.rData` files, otherwise saved values will be used. The `tree.sample.size` variable is set at 400, which is an acceptable value for the full test, but this is slow and it can be set to a lower value for testing.

```{r echo=TRUE}
library(logisticPCA)
library(ape)
library(vegan)
library(parallel)
library(phytools)
library(rnaturalearth)
library(tidyverse)
library(ggrepel)
library(gridExtra) # for grid.arrange

MC.CORES <- detectCores() # Use all available processing power
cat(paste("--> Using", MC.CORES, "cores\n"))
```

As a sanity check, transform the pronoun paradigm similarities into an inter-language distance measure. A cluster plot of the distance matrix showing structurally similar languages clustered together. Note that this is *not* a phylogeny!

```{r}
# The similarity features are comparisions of every pair of cells in a pronoun paradigm, 
# calculated per language
pairwise.similarity <- read.delim("pronoun-similarity.csv", row.names=1, header=TRUE)
language.distance <- dist(pairwise.similarity,  method="manhattan") / ncol(pairwise.similarity)
plot(hclust(language.distance), cex=0.5)
```

Prepare metadata 
```{r}
metadata <- read_tsv("language-metadata.csv")
sae.features <- read.delim("SAE-features.csv", row.names=1, na.string="NA")
sae.sum <- apply(sae.features, 1, sum, na.rm=TRUE)[metadata$language]
metadata$sae.score <- sae.sum
metadata$SAE <- sae.sum > 4
metadata
#View(metadata)
```


The pairwise similarity measures are summarised using Logistic PCA (designed for binary data) into two continuous components.

```{r}
#pairwise.similarity <- read.delim("pronoun-similarity.csv", row.names=1)
logpca_cv = cv.lpca(pairwise.similarity, ks = 2, ms = 1:10)
plot(logpca_cv)
```
```{r}
which.min(logpca_cv)
```



```{r}
logpca_model = logisticPCA(pairwise.similarity, k = 2, m = which.min(logpca_cv))
#plot(logpca_model, type = "scores")
logpca_model$PCs %>% 
  as.data.frame() %>%
  mutate(language=rownames(.)) %>%
  rename(PC1=V1, PC2=V2) %>%
  left_join(metadata) %>%
 ggplot(aes(x=PC1, y=PC2, label=language, colour=SAE)) + 
 geom_point() + 
 scale_colour_manual(values=c("darkgray", "black")) +
 geom_text_repel(size=3, segment.size=0.25, segment.alpha=0.5, box.padding=0.1) +
 theme(legend.position = "none")
```
```{r}
ggsave("../figures/Figure_2_PCA.png", width=5.75, height=5.75, units="in")
ggsave("../figures/Figure_2_PCA.pdf", width=5.75, height=5.75, units="in")
```


```{r}
logpca_model$PCs %>% 
  as.data.frame() %>%
  mutate(language=rownames(.)) %>%
  rename(PC1=V1, PC2=V2) %>%
  left_join(read_tsv("pronoun-number-conflation.csv")) %>% 
  #left_join(metadata) %>%
  ggplot(aes(x=PC1, y=PC2, label=language, colour=as.factor(conflation))) + 
  geom_point() + #geom_point(aes(shape=SAE)) + 
  geom_text_repel(size=3, segment.size=0.25, segment.alpha=0.5, box.padding=0.1) +
  labs(colour="Conflated\nnumber")
```


```{r}
logpca_model$U %>% 
  as_tibble() %>% 
  mutate(features=names(pairwise.similarity)) %>% 
  rename(PC1=V1, PC2=V2) %>%
  mutate(magnitude=sqrt(PC1^2 + PC2^2)) %>%
  ggplot(aes(x=PC1, y=PC2)) + 
  geom_point() + 
  geom_text_repel(aes(label=if_else(magnitude > 0.1, features, "")), size=3, segment.size=0.25, segment.alpha=0.5, box.padding=0.1)
```

## Data transformation

Prepare the tree sample if the file `trees.rData` doesn't already exist (this is a slow process, so we don't want to repeat it if not necessary). To force regeneration of `trees.rData` delete the current version of the file along with other `.rData` files.

Sanity check: Load a reference tree, fix the tree labels, and plot it
```{r}
all.languages <- metadata %>% pull(language)
namehash <- c(
        "Albanian_G"="Albanian_Gheg",
        "Albanian_Standard"="Albanian_Tosk",
        "Armenian_Mod"="Armenian",
        "Breton_ST"="Breton",
        "Bihari"="Maithili",
        "Dutch_List"="Dutch",
        "German_Munich"="German",
        "Greek_Mod"="Modern_Greek",
        "Icelandic_ST"="Icelandic",
        "Lithuanian_ST"="Lithuanian",
        "Luxembourgish"="Luxemburgish",
        "Portuguese_ST"="Portuguese",
        "Rumanian_List"="Romanian",
        "Sardinian_C"="Sardinian_Campidanese",
        "Sardinian_L"="Sardinian_Logudorese",
        "Singhalese"="Sinhalese",
        "Tadzik"="Tajik")

read.nexus("dollo.mcc.tre")
t <- read.nexus("dollo.mcc.tre")
t$tip.label <- recode(t$tip.label, !!!namehash)
dollo.mcc.tree <- ladderize(root(keep.tip(t, all.languages), "Hittite"))
plot(dollo.mcc.tree, no.margin=TRUE, cex=0.6)
```

```{r}
logisticPC1 <- logpca_model$PCs[,1]
logisticPC2 <- logpca_model$PCs[,2]

states.pca <- data.frame(
  PC1=round(logisticPC1,5), 
  PC2=round(logisticPC2,5))

tree.sample.size <- 400 # set this to e.g. 20 for testing, 400+ for a full analysis

if (!file.exists("trees.rData")){
  TREEFILE <- "dollo.combined.trees" # this sample is 3600 trees, which is too long
  
  trees <- read.nexus(TREEFILE)
  if (length(trees) > tree.sample.size){
    trees <- trees[seq(1, length(trees), length.out=tree.sample.size)]
  }

  tree.list <- mclapply(trees, 
    function(tree) {
      # fix labels
      tree$tip.label <- recode(tree$tip.label, !!!namehash)
      # normalise edge lengths (there are sometimes errors when numbers are too large)
      tree$edge.length <- tree$edge.length / 1000
      # root on Hittite
      ladderize(root(keep.tip(tree, all.languages), "Hittite"))
      }, 
    mc.cores=MC.CORES)

  class(tree.list) <- "multiPhylo"
  
  save(tree.list, states.pca, dollo.mcc.tree, file="trees.rData")
}
```


## Simulate SAE history

Plot these states on a map. 

```{r}
world = ne_countries(scale = "small", returnclass = "sf")
metadata <- metadata %>% mutate(label=paste(iso.code, sae.score))

ggplot(data=world) + 
  geom_sf(fill="white") +
  coord_sf(xlim = range(metadata$longitude), ylim = range(metadata$latitude)) +
  geom_label(data=metadata, aes(x = longitude, y = latitude, label=label, colour=ancient, fill=SAE), size=3, label.padding = unit(0.15, "lines")) +
  scale_fill_manual(values=c("lightgray","darkgray")) +
  scale_colour_manual(values=c("black", "white")) +
  theme(legend.position = "none", axis.title.x=element_blank(), axis.title.y=element_blank())
```

```{r}
ggsave("../figures/Figure_1_Map.pdf", width=5.75, height=3.75, units="in")
ggsave("../figures/Figure_1_Map.png", width=5.75, height=3.75, units="in")
```


```{r}
metadata %>% 
  ggplot(aes(x=sae.score)) + 
  geom_histogram(binwidth=1) + 
  scale_x_continuous(breaks=0:9) +
  xlab("SAE Features") + ylab("Count")
```
Natural breakpoint at sae.score >= 5
```{r}
ggsave("../figures/Figure_3_SAE.png", width=3, height=2)
ggsave("../figures/Figure_3_SAE.pdf", width=3, height=2)
```

## Demonstrate a simmap
```{r}
observed.states <- metadata %>% pull(SAE) %>% as.integer()
names(observed.states) <- metadata %>% pull(language)
# make.simmap rate models: "ARD" all rates different; "SYM" symmetrical; "EQ"" equal; or use matrix
dollo.model = matrix(c(0,0,1,0), 2) # once a lineage becomes SAE it doesn't revert to non-SAE
simmap.tre <- make.simmap(dollo.mcc.tree,
  observed.states,
  model=dollo.model)
cols <- c("black", "gray")
names(cols) <- c("0", "1")
plotSimmap(simmap.tre, cols, fsize=0.5, pts=FALSE, lwd=4)
```

Create all the simmaps. This can be an extremely slow process
```{r include=FALSE}
if (!file.exists("simmap.rData")){
  simmap.list <- make.simmap(tree.list, observed.states, model=dollo.model, nsim=5)
  save(simmap.list, file="simmap.rData")
} else load("simmap.rData")
```

Plot three arbitrary simmaps to show variation in tree structure (from the Bayesian phylogenetic sampline process) and simulated history
```{r}
#library(ggplotify)
opar <- par(no.readonly=TRUE)
par(mfrow=c(1,3))
plot.simmap.i <- function(i){
  cols <- c("black", "gray")
  names(cols) <- c("0", "1")
  plot(simmap.list[[i]], cols, fsize=0.5, pts=FALSE, lwd=4)}

plot.simmap.i(1)
plot.simmap.i(2)
plot.simmap.i(3)
par(opar)
```
```{r}
pdf("../figures/Figure_4_State_Simulation.pdf", width=7, height=5.5)
par(mfrow=c(1,3))
plot.simmap.i(1)
plot.simmap.i(2)
plot.simmap.i(3)
dev.off()
```



Now do the `brownie.lite` test. The standard citation for this is:

  O'Meara, B.C., C. Ane, M.J. Sanderson, and P.C. Wainwright. 2006. Testing for different rates of continuous trait evolution using likelihood. *Evolution* 60(5): 922-933.


```{r}
#states <- cbind(states.mds, states.pca)
#states <- states.pca
NTREES <- length(simmap.list) 
results.trait.m <- matrix( # template results matrix
  nrow=NTREES,
  ncol=12, 
  dimnames=list(1:NTREES,
    c("BM.singlerate","logL.singlerate", "k1", "1rate.AICc", "NonSAE.rate",
      "SAE.rate","logL.multiplerate", "k2", "2rate.AICc", "pchisq",
     "converged", "delta.AICc"))
  )

if (!file.exists("results.rData")){
  #results <- list()
  process.results <- function(trait.name){
    results.trait.name <- results.trait.m
    traits <- states.pca[[trait.name]]
    names(traits) <- row.names(states.pca)
    ## traits <- traits[simmap.list[[1]]$tip.label] 
    for (i in 1:NTREES){
      cat(paste("Trait ", trait.name, " trial ", i, "/", NTREES, "\n", sep=""))
      # check that labels are aligned
      stopifnot(all(sort(simmap.list[[i]]$tip.label) == sort(names(traits))))
      fit <- NA
      tryCatch(
        fit <- brownie.lite(simmap.list[[i]], traits, test="chisq"),
          error=function(e) {}
        )
        if (length(fit) != 1){
          # calculate AICc for 1 and 2 rate models
          fit1aicc<-(-2*fit$logL1+2*fit$k1+((2*fit$k1*(fit$k1+1))/(90-fit$k1-1))) 
          fit2aicc<-(-2*fit$logL.multiple+2*fit$k2+((2*fit$k2*(fit$k2+1))/(90-fit$k2-1))) 

          results.trait.name[i,]<- as.numeric(c(
              fit$sig2.single, 
              fit$logL1, 
              fit$k1, 
              fit1aicc, 
              fit$sig2.multiple[sort(names(fit$sig2.multiple))][[1]],
              fit$sig2.multiple[sort(names(fit$sig2.multiple))][[2]],
              fit$logL.multiple,
              fit$k2,
              fit2aicc,
              fit$P.chisq,
              fit$convergence == "Optimization has converged.",
              fit1aicc-fit2aicc
              )) 
        }
    }
    results.trait.name
  }
  results <- mclapply(names(states.pca), process.results,
    mc.cores=min(MC.CORES, length(names(states.pca))))
  names(results) <- names(states.pca)
  save(results, file="results.rData")
} else load("results.rData")
```

Get rate estimates (report in article text)
```{r}
trait_labels <- c("PC1", "PC2")
lapply(trait_labels, function(d) as_tibble(results[[d]])) %>%
  set_names(trait_labels) %>%
  bind_rows(.id="trait") %>% 
  pivot_longer(c("logL.singlerate", "logL.multiplerate"), names_to="model", values_to="rate") %>%
  group_by(trait, model) %>%
  summarise(median_rate=median(rate))
```

## delta AICc

Compare AIC for one rate and two rate models

```{r}
trait_labels <- c("PC1", "PC2")

lapply(trait_labels, function(d) as_tibble(results[[d]])) %>%
  set_names(trait_labels) %>%
  bind_rows(.id="trait") %>% 
  pivot_longer(c("1rate.AICc", "2rate.AICc"), names_to="model", values_to="AICc") %>%
  ggplot(aes(x=trait, y=AICc, colour=model)) + geom_boxplot()
```

Change in AIC gives a measure of how much one model should be preferred over the other
```{r}
fig5c <- lapply(trait_labels, function(d) as_tibble(results[[d]])) %>%
  set_names(trait_labels) %>%
  bind_rows(.id="trait") %>% 
  ggplot(aes(x=trait, y=delta.AICc)) + 
  geom_boxplot() +
  labs(title="(c)") +
  theme(axis.title.x=element_blank())
fig5c
```

Report median change in AICc in text
```{r}
lapply(trait_labels, function(d) as_tibble(results[[d]])) %>%
  set_names(trait_labels) %>%
  bind_rows(.id="trait") %>%
  group_by(trait) %>%
  summarise(median.delta.AIC.c=median(delta.AICc))
```

For the two rate model, how the estimated rates of evolution of each component in the SAE and non-SAE regions of the trees.

```{r}
fig5a <- lapply(trait_labels, function(d) as_tibble(results[[d]])) %>%
  set_names(trait_labels) %>%
  bind_rows(.id="trait" ) %>%
  pivot_longer(c("SAE.rate", "NonSAE.rate"), names_to="area", values_to="rate") %>%
  mutate(area=if_else(area=="SAE.rate", "SAE", "non-SAE")) %>%
  ggplot(aes(x=trait, y=rate, fill=area)) + 
  geom_boxplot() + 
  theme(legend.position = "bottom") + 
  scale_fill_grey(start=0.5, end=0.8)+
  labs(title="(a)") +
  theme(axis.title.x=element_blank())
fig5a
```

Summarise the relative rate of evolution of each component for SAE versus non-SAE parts of the tree
```{r}
fig5b <- lapply(trait_labels, function(d) as_tibble(results[[d]])) %>%
  set_names(trait_labels) %>%
  bind_rows(.id="trait" ) %>%
  mutate(rate.multiplier=SAE.rate/NonSAE.rate) %>%
  ggplot(aes(x=trait, y=rate.multiplier)) + geom_boxplot() +
  labs(title="(b)") +
  theme(axis.title.x=element_blank())
fig5b
```
```{r}
lapply(trait_labels, function(d) as_tibble(results[[d]])) %>%
  set_names(trait_labels) %>%
  bind_rows(.id="trait" ) %>%
  mutate(rate.multiplier=SAE.rate/NonSAE.rate) %>%
  group_by(trait) %>%
  summarise(median.rate.multiplier=median(rate.multiplier))
```

```{r}
#fig5 <- grid_arrange_shared_legend(fig5a, fig5b, fig5c, nrow=1, position="right")
fig5 <- grid.arrange(fig5a, fig5b, fig5c, nrow=1)
```

```{r}
ggsave("../figures/Figure_5_results.png", plot=fig5, width=7, height=3)
ggsave("../figures/Figure_5_results.pdf", plot=fig5, width=7, height=3)
```

